Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-11-19 revision/r2/response_r2.tex

% \begin{mybox}
% \emph{I would love to actually read the authors’ thoughts here and ideally be proven wrong that an
% improved workflow would be enough for enabling policy decisions whose beneficial impacts
% would be commensurate with the current magnitude of biodiversity loss7. In particular, I
% would be interested in a better articulation of how they see global studies feed-backing on
% policy which is usually more local, national or regional.}  
% \end{mybox}

% ...

\begin{mybox}
\emph{My remaining points are more concise.
- Terminology. If the authors are also aiming at the policy sphere, the word bias (e.g. lines
13, 134) should be wielded with extreme caution as it is easily weaponized. In policy
circles, the term bias is, in my experience, the kiss of death as it is often interpreted as
lopsided evidence. The authors are discussing preferential sampling with respect to space,
time or taxonomy and its consequences in modelling results. Preferential sampling is more
descriptive than bias, which also has a prescriptive meaning, esp. in policy. Also, a well-
designed survey at some spatial scale may become an instance of preferential geographic
sampling when collated with other surveys and data. This sort of data collation is often the
case with global trend analyses that the authors seem to have in mind.}  
\end{mybox}

This is a good point, we have clarified our use of this word in the two areas we used it without additional context (lines, \href{file:forecastflows_r2\#lintarget:bias1}{\ref*{lin:bias1}}, \href{file:forecastflows_r2\#lintarget:bias2}{\ref*{lin:bias2}}). % emw18Oct: fix these refs to main text!
% Hmm... well... I prefer bias. Preferential sampling sounds to me like it's done on purpose

### 2025-11-19 revision/r2/forecastflows_r2.tex
fix ref to main text

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated), flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see Supplementary Figure 1) or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased (for example, more intensive sampling in Europe) \llabel{bias2}, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.

>>>>

A key feature of this workflow is the central role of data simulation, which introduces two feedback loops.
The first feedback arises when we evaluate the model on simulated data.
The failure of the model to recover known parameter values and handle the complexity of the simulated data should prompt reconsidering the model, or even reformulating the research question.
Further, this step might reveal that some parameters are highly non-identifiable (meaning the parameter(s) cannot be uniquely estimated), flagging the need to change the model structure---before incorporating empirical observations. 
The second feedback loop comes from the retrodictive check. 
Discrepancies here may indicate a missing key driver, and suggest the current model is too simplistic. We can refine the model to integrate the missing process(es)---if we can identify them---and return to the start of the workflow. Insights from the retrodictive check can also lead us to introduce additional complexity when simulating fake data, such as phylogenetic structure \llabel{suppfig2}(see Supplementary Figure 1) or observational biases (e.g. unbalanced data). For example, if a researcher realizes their empirical data is geographically biased (for example, more intensive sampling in Europe)\llabel{bias2}, this bias should be built into the model and thus into this data simulation step. This iterative evaluation of the model moves beyond a simple reliance on goodness-of-fit metrics. At each iteration, we are able to evaluate the model behavior, both with simulated and real data, taking into account our expert knowledge of the ecological processes. \llabel{broadcontext}These feedback loops are applied currently only in Bayesian frameworks (to our knowledge), but could and should extend across other modeling approaches given their power to improve models and, in turn, scientific understanding.
